The Radio Sky at Meter Wavelengths: $m$-Mode Analysis Imaging\\ with the OVRO-LWA

### 2018-07-06 paper.tex
remove revision markup

\title{
    The Radio Sky at Meter Wavelengths: $m$-Mode Analysis Imaging\\
    with the \revision{OVRO-LWA}
}

@@

Just as for CMB experiments, foreground removal or suppression is an essential component of both
classes of 21-cm cosmology experiments. The brightness temperature of the galactic synchrotron
emission at high galactic latitudes is measured by \revision{\citet{2017MNRAS.464.4995M}} as
\revision{\begin{equation}
    T \sim 300\,{\rm K} \times \left(\frac{\nu}{150\,{\rm MHz}}\right)^{-2.6}\,.
\end{equation}}
Therefore, experiments conservatively need to achieve 5 orders of dynamic range against this
foreground emission before the cosmological signal can be measured. Current foreground removal
methods (for example, \citealt{2012ApJ...756..165P} and \citealt{2013MNRAS.429..165C}) rely on the
assumption that the foreground emission is spectrally smooth. The low-frequency radio sky is
composed of several components: galactic synchrotron emission, supernova remnants, radio galaxies,
free-free emission and absorption from \ion{H}{2} regions, and a confusing background of radio
sources.  Ideally, a foreground removal strategy should be informed by the measured spatial
structure and frequency spectrum of all foreground components. For instance, CMB experiments
typically construct several maps at several frequencies to enable component separation.  At low
frequencies, this possibility is limited by the availability of suitable high-fidelity sky maps on
angular scales ranging from tens of degrees to arcminutes.

@@

As a demonstration of this technique we apply Tikhonov regularized $m$-mode analysis imaging to the
Owens Valley \revision{Radio Observatory} Long Wavelength Array (OVRO-LWA), and generate a series of new
low-frequency maps of the sky between 36.528~MHz and 73.152~MHz.  These maps capture the full sky
visible from the Owens Valley Radio Observatory (OVRO) with angular resolution of $\sim 15$~arcmin.
These new maps complement the existing full-sky maps at these frequencies with greatly improved
angular resolution.

@@

The structure of this paper is as follows. In \S\ref{sec:imaging}, we present Tikhonov-regularized
$m$-mode analysis imaging, a new imaging technique that allows us to image the entire visible sky in
one coherent synthesis imaging step with exact widefield corrections. In \S\ref{sec:observations} we
describe our observations with the \revision{OVRO-LWA}. In \S\ref{sec:results} we present the sky
maps and compare these maps against other low-frequency sky maps.  In \S\ref{sec:error} we discuss
some of the sources of error present in the maps, and finally in \S\ref{sec:conclusion} we present
our conclusions.

@@

The key advantage of $m$-mode analysis is the block-diagonal structure of
Equation~\ref{eq:m-mode-matrix-equation}. The computational complexity of many common matrix
operations (eg. solving a linear system of equations) is $\mathcal{O}(N^3)$, \revision{where $N$ is
the linear size of the matrix.  By splitting the equation into $M$ independent blocks, the number of
floating point operations required to solve the linear system of equations is now
$\mathcal{O}(N^3M^{-2})$, because each block can be manipulated independently of the other blocks.
This computational savings is what makes this matrix algebra approach to interferometric imaging
feasible. For the dataset presented in this paper, computing the elements of the transfer matrix
takes $\sim$10 hours per frequency channel on a 10-node cluster, but once the matrix has been
computed the imaging process described in \S\ref{sec:mmode-imaging} takes $\sim$10 minutes, and the
deconvolution process described in \S\ref{sec:clean} was allowed to run for $\sim$10 hours.}

@@

\subsection{The Owens Valley \revision{Radio Observatory} Long Wavelength Array}

@@

The Owens Valley \revision{Radio Observatory} Long Wavelength Array (OVRO-LWA) is a 288-element
interferometer located at the Owens Valley Radio Observatory near Big Pine, California
\citep{hallinan_2017}.  The OVRO-LWA is a low-frequency instrument with instantaneous bandwidth
covering 27~MHz to 85~MHz and with 24~kHz channelization.  Each antenna stand hosts two
perpendicular broadband dipoles so that there are $288\times2$ signal paths in total. These signal
paths feed into the 512-input LEDA correlator \citep{2015JAI.....450003K}, which allows the OVRO-LWA
to capture the entire visible hemisphere in a single snapshot image.

@@

\revision{
    To achieve this dynamic range at low frequencies, it is important to account for propagation
    effects through the ionosphere. In the weak scattering regime ($r_\text{diff} \gg r_\text{f}$,
    where $r_\text{diff}$ is the diffractive scale of the ionosphere, $r_\text{f} =
    \sqrt{\lambda D / 2\pi}$ is the Fresnel scale, $\lambda$ is the wavelength, and $D$ is the
    distance to the ionosphere), fluctuations within the ionosphere contribute
    amplitude and phase scintillations that can be described by a direction-dependent complex-gain
    calibration. This justifies the use of ``peeling'', which incorporates a direction-dependent
    calibration to subtract sources in the presence of ionospheric scintillation.
    \citep[eg.][]{2008ISTSP...2..707M, 2015MNRAS.449.2668S}.
}

\revision{
    In the strong scattering regime ($r_\text{diff} \lesssim r_\text{f}$), the image of a point
    source can ``break apart'' into multiple images or speckles \citep{2015MNRAS.453..925V}.
    Attempting to peel a source in the strong scattering regime will lead to source subtraction
    artifacts in the final sky map.  \citet{2016RaSc...51..927M} measured that from the location of
    LOFAR at 150 MHz, the diffractive scale of the ionosphere is $>5$ km 90\% of the time. This
    implies that at 73 MHz, the diffractive scale is typically $>2$ km, and at 36 MHz, the
    diffractive scale is typically $>1$ km. These limits are comparable to the Fresnel scale for the
    OVRO-LWA (ie. $r_\text{diff} > r_\text{f}$), and therefore we do not generally expect to see
    strong scattering from the ionosphere.  Ionospheric conditions during the observing period were
    mild (see \S\ref{sec:ionosphere-conditions}). However, we do observe scintillation and
    refractive-offset events on the timescale of a single integration (13 seconds, see
    Figure~\ref{fig:scintillation}). Consequently we peeled Cyg~A and Cas~A from the dataset using a
    new solution for each integration.
}

\revision{
    In addition, the largest angular scale of Cas~A is $\sim$8~arcmin, and the largest angular scale
    of Cyg~A is $\sim$2~arcmin. With a $\sim10$~arcmin resolution on its longest baselines at
    73~MHz, the OVRO-LWA marginally resolves both sources. A resolved source model is needed for
    both sources. We fit a self-consistent resolved source model to each source. This is performed
    by minimizing the variance within an aperture located on each source after peeling. By phasing
    up a large number of integrations before imaging (at least 1 hour), it is possible to smear out
    the contribution of the rest of the sky.  We then use a non-linear optimization routine (NLopt
    Sbplx) \citep{nlopt, sbplx} to vary the parameters in a source model until the variance within
    the aperture is minimized. Cyg~A is modeled with 2 Gaussian components, while Cas~A is modeled
    with 3~Gaussian components.  Ultimately, these multi-component models are used to peel Cyg~A and
    Cas~A, but residual errors from this model and from the ionosphere (particularly while these
    sources are at low elevations) contribute residual artifacts that are largely localized to
    within 1$^\circ$ of each source.
}

@@

The noise in each map is empirically measured using jackknife resampling. The dataset is first split
into even- and odd-numbered integrations. These two groups are then imaged and CLEANed
independently, before being compared against the maps constructed from all of the available data
using the jackknife \revision{standard error} estimator. This estimate of the \revision{standard
error} includes all sources of error that operate on $\sim13$ second timescales (the integration
time) such as thermal noise and rapid ionospheric fluctuations, but does not account for more slowly
varying effects (for example, sidereal variation in the system temperature or day-night fluctuations
in the ionosphere).  These noise calculations are summarized in Table~\ref{tab:summary}.  VLSSr
source counts \citep{2014MNRAS.440..327L} suggest that the confusion limit at 74 MHz and 15 arcmin
angular resolution is $\sim 1000\times(\nu/74\,{\rm MHz})^{-0.7}$ mJy.  Each channel map achieves
thermal noise $<900$ mJy and therefore each map is likely at or near the confusion limit.

@@

The LWA1 Low Frequency Sky Survey (LLFSS) \citep{2017MNRAS.469.4537D} produced nine maps of the sky
between 35~MHz and 80~MHz. Six of these maps are interior to the frequency range spanned by this
work. \revision{Initial comparisons with the LLFSS helped characterize a systematic rotation in the
LWA1's antenna positions. After phase-calibration, this manifested itself as a systematic rotation
and translation in the snapshot images that were mosaicked to form the final sky map. This
systematic error has been corrected in the comparisons presented here and in the latest version of
the LLFSS\footnote{
    Available for download at: \url{http://lda10g.alliance.unm.edu/LWA1LowFrequencySkySurvey/}
}.}

\revision{A direct comparison with these updated LLFSS maps can be seen in
Figure~\ref{fig:lwa1-comparison}.  In this figure, the LLFSS maps are filtered to remove the
monopole and all modes with $m=0$. The OVRO-LWA maps are interpolated in frequency and blurred to
match the angular resolution of the corresponding LLFSS map.  At 60~MHz, the agreement is generally
better than 10\%. However at lower frequencies the agreement deteriorates to about 20\%.  Typically
the OVRO-LWA maps have excess emission in the galactic plane and a deficit of emission off the
galactic plane relative to the LLFSS.}

>>>>

\title{
    The Radio Sky at Meter Wavelengths: $m$-Mode Analysis Imaging\\
    with the OVRO-LWA
}

@@

Just as for CMB experiments, foreground removal or suppression is an essential component of both
classes of 21-cm cosmology experiments. The brightness temperature of the galactic synchrotron
emission at high galactic latitudes is measured by \citet{2017MNRAS.464.4995M} as
\begin{equation}
    T \sim 300\,{\rm K} \times \left(\frac{\nu}{150\,{\rm MHz}}\right)^{-2.6}\,.
\end{equation}
Therefore, experiments conservatively need to achieve 5 orders of dynamic range against this
foreground emission before the cosmological signal can be measured. Current foreground removal
methods (for example, \citealt{2012ApJ...756..165P} and \citealt{2013MNRAS.429..165C}) rely on the
assumption that the foreground emission is spectrally smooth. The low-frequency radio sky is
composed of several components: galactic synchrotron emission, supernova remnants, radio galaxies,
free-free emission and absorption from \ion{H}{2} regions, and a confusing background of radio
sources.  Ideally, a foreground removal strategy should be informed by the measured spatial
structure and frequency spectrum of all foreground components. For instance, CMB experiments
typically construct several maps at several frequencies to enable component separation.  At low
frequencies, this possibility is limited by the availability of suitable high-fidelity sky maps on
angular scales ranging from tens of degrees to arcminutes.

@@

As a demonstration of this technique we apply Tikhonov regularized $m$-mode analysis imaging to the
Owens Valley Radio Observatory Long Wavelength Array (OVRO-LWA), and generate a series of new
low-frequency maps of the sky between 36.528~MHz and 73.152~MHz.  These maps capture the full sky
visible from the Owens Valley Radio Observatory (OVRO) with angular resolution of $\sim 15$~arcmin.
These new maps complement the existing full-sky maps at these frequencies with greatly improved
angular resolution.

@@

The structure of this paper is as follows. In \S\ref{sec:imaging}, we present Tikhonov-regularized
$m$-mode analysis imaging, a new imaging technique that allows us to image the entire visible sky in
one coherent synthesis imaging step with exact widefield corrections. In \S\ref{sec:observations} we
describe our observations with the OVRO-LWA. In \S\ref{sec:results} we present the sky maps and
compare these maps against other low-frequency sky maps.  In \S\ref{sec:error} we discuss some of
the sources of error present in the maps, and finally in \S\ref{sec:conclusion} we present our
conclusions.

@@

The key advantage of $m$-mode analysis is the block-diagonal structure of
Equation~\ref{eq:m-mode-matrix-equation}. The computational complexity of many common matrix
operations (eg. solving a linear system of equations) is $\mathcal{O}(N^3)$, where $N$ is the linear
size of the matrix.  By splitting the equation into $M$ independent blocks, the number of floating
point operations required to solve the linear system of equations is now $\mathcal{O}(N^3M^{-2})$,
because each block can be manipulated independently of the other blocks.  This computational savings
is what makes this matrix algebra approach to interferometric imaging feasible. For the dataset
presented in this paper, computing the elements of the transfer matrix takes $\sim$10 hours per
frequency channel on a 10-node cluster, but once the matrix has been computed the imaging process
described in \S\ref{sec:mmode-imaging} takes $\sim$10 minutes, and the deconvolution process
described in \S\ref{sec:clean} was allowed to run for $\sim$10 hours.

@@

\subsection{The Owens Valley Radio Observatory Long Wavelength Array}

@@

The Owens Valley Radio Observatory Long Wavelength Array (OVRO-LWA) is a 288-element interferometer
located at the Owens Valley Radio Observatory near Big Pine, California \citep{hallinan_2017}.  The
OVRO-LWA is a low-frequency instrument with instantaneous bandwidth covering 27~MHz to 85~MHz and
with 24~kHz channelization.  Each antenna stand hosts two perpendicular broadband dipoles so that
there are $288\times2$ signal paths in total. These signal paths feed into the 512-input LEDA
correlator \citep{2015JAI.....450003K}, which allows the OVRO-LWA to capture the entire visible
hemisphere in a single snapshot image.

@@

To achieve this dynamic range at low frequencies, it is important to account for propagation effects
through the ionosphere. In the weak scattering regime ($r_\text{diff} \gg r_\text{f}$, where
$r_\text{diff}$ is the diffractive scale of the ionosphere, $r_\text{f} = \sqrt{\lambda D / 2\pi}$
is the Fresnel scale, $\lambda$ is the wavelength, and $D$ is the distance to the ionosphere),
fluctuations within the ionosphere contribute amplitude and phase scintillations that can be
described by a direction-dependent complex-gain calibration. This justifies the use of ``peeling'',
which incorporates a direction-dependent calibration to subtract sources in the presence of
ionospheric scintillation.  \citep[eg.][]{2008ISTSP...2..707M, 2015MNRAS.449.2668S}.

In the strong scattering regime ($r_\text{diff} \lesssim r_\text{f}$), the image of a point source
can ``break apart'' into multiple images or speckles \citep{2015MNRAS.453..925V}.  Attempting to
peel a source in the strong scattering regime will lead to source subtraction artifacts in the final
sky map.  \citet{2016RaSc...51..927M} measured that from the location of LOFAR at 150 MHz, the
diffractive scale of the ionosphere is $>5$ km 90\% of the time. This implies that at 73 MHz, the
diffractive scale is typically $>2$ km, and at 36 MHz, the diffractive scale is typically $>1$ km.
These limits are comparable to the Fresnel scale for the OVRO-LWA (ie. $r_\text{diff} >
r_\text{f}$), and therefore we do not generally expect to see strong scattering from the ionosphere.
Ionospheric conditions during the observing period were mild (see
\S\ref{sec:ionosphere-conditions}). However, we do observe scintillation and refractive-offset
events on the timescale of a single integration (13 seconds, see Figure~\ref{fig:scintillation}).
Consequently we peeled Cyg~A and Cas~A from the dataset using a new solution for each integration.

In addition, the largest angular scale of Cas~A is $\sim$8~arcmin, and the largest angular scale of
Cyg~A is $\sim$2~arcmin. With a $\sim10$~arcmin resolution on its longest baselines at 73~MHz, the
OVRO-LWA marginally resolves both sources. A resolved source model is needed for both sources. We
fit a self-consistent resolved source model to each source. This is performed by minimizing the
variance within an aperture located on each source after peeling. By phasing up a large number of
integrations before imaging (at least 1 hour), it is possible to smear out the contribution of the
rest of the sky.  We then use a non-linear optimization routine (NLopt Sbplx) \citep{nlopt, sbplx}
to vary the parameters in a source model until the variance within the aperture is minimized. Cyg~A
is modeled with 2 Gaussian components, while Cas~A is modeled with 3~Gaussian components.
Ultimately, these multi-component models are used to peel Cyg~A and Cas~A, but residual errors from
this model and from the ionosphere (particularly while these sources are at low elevations)
contribute residual artifacts that are largely localized to within 1$^\circ$ of each source.

@@

The noise in each map is empirically measured using jackknife resampling. The dataset is first split
into even- and odd-numbered integrations. These two groups are then imaged and CLEANed
independently, before being compared against the maps constructed from all of the available data
using the jackknife standard error estimator. This estimate of the standard error includes all
sources of error that operate on $\sim13$ second timescales (the integration time) such as thermal
noise and rapid ionospheric fluctuations, but does not account for more slowly varying effects (for
example, sidereal variation in the system temperature or day-night fluctuations in the ionosphere).
These noise calculations are summarized in Table~\ref{tab:summary}.  VLSSr source counts
\citep{2014MNRAS.440..327L} suggest that the confusion limit at 74 MHz and 15 arcmin angular
resolution is $\sim 1000\times(\nu/74\,{\rm MHz})^{-0.7}$ mJy.  Each channel map achieves thermal
noise $<900$ mJy and therefore each map is likely at or near the confusion limit.

@@

The LWA1 Low Frequency Sky Survey (LLFSS) \citep{2017MNRAS.469.4537D} produced nine maps of the sky
between 35~MHz and 80~MHz. Six of these maps are interior to the frequency range spanned by this
work. Initial comparisons with the LLFSS helped characterize a systematic rotation in the LWA1's
antenna positions. After phase-calibration, this manifested itself as a systematic rotation and
translation in the snapshot images that were mosaicked to form the final sky map. This systematic
error has been corrected in the comparisons presented here and in the latest version of the
LLFSS\footnote{
    Available for download at: \url{http://lda10g.alliance.unm.edu/LWA1LowFrequencySkySurvey/}
}.

A direct comparison with these updated LLFSS maps can be seen in Figure~\ref{fig:lwa1-comparison}.
In this figure, the LLFSS maps are filtered to remove the monopole and all modes with $m=0$. The
OVRO-LWA maps are interpolated in frequency and blurred to match the angular resolution of the
corresponding LLFSS map.  At 60~MHz, the agreement is generally better than 10\%. However at lower
frequencies the agreement deteriorates to about 20\%.  Typically the OVRO-LWA maps have excess
emission in the galactic plane and a deficit of emission off the galactic plane relative to the
LLFSS.
